---
title: "  "
output: 
  bookdown::html_document2:
    includes:
      after_body: assets/foot.html
---

```{r setup, include=FALSE, purl=FALSE, message=FALSE}

knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(icon)

```

<br>
<br>

:::obj

 **Objectives for this section:**
 
 - look at a messy spreadsheet and think about ways we could make it tidy
 - read in data from different spreadsheet tabs
 - write data to a spreadsheet tab and a CSV!
 
:::

<br>
`r icon::fa("r-project", color="steelblue")`
<br>

## How to get data from a spreadsheet into R!

Spreadsheets surround us and exist in nearly every facet and field. While learning R will make life easier and more reproducible, spreadsheets won't go away, and there will always still be a messy dataset that comes your way.


<details>
  <summary>Click for Answer</summary>

So here's an example of a plot with more info on something. We can use the binoculars or magnifying glass to signify "*more info here if you want*"

    ```{r, echo=FALSE, eval=TRUE}
    hist(1:10)
    ```

</details> 
### What is Tidy Data?


```{r, echo=TRUE, eval=TRUE}
library(tidyverse)
library(openxlsx)
library(readxl)

# load data
messy <- readxl::read_excel(path = "data/csci_spreadsheet.xlsx", sheet = 1)
```

